---
title: "Intro to R and R- Spatial; insight on datacubes"
author: "Edzer Pebesma"
date: "2/16/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (Sys.getenv("USER") == "edzer") {
		set.seed(15751)
}
```

## What is R?

According to the [website](https://r-project.org), "R is a free
software environment for statistical computing and graphics. It
compiles and runs on a wide variety of UNIX platforms, Windows
and MacOS."


## What is Rstudio?

[Rstudio](https://rstudio.com/) is a company that develops and
distributes a frontend, or IDE, for R that is free and open source,
called Rstudio desktop. It also develops commercial derivatives,
mostly targeted at businesses. These products use the same R
(processor) as R.

Other front-ends include 

* R command line
* teminal-based, e.g. vim with Nvim-R, or emacs with ESS
* a browser communicating with Rstudio server (also free)
* R in visual studio
* Jupyter or JupyterLab (Jupyter developed from IPython notebooks and once stood for "Julia, Python, R", it now probably caters more languages)

## What is markdown, R-markdown?

[Markdown](https://en.wikipedia.org/wiki/Markdown) is a leightweight
markup language for formatted text (headers, lists, figures, equations). 

[R-markdown](https://rmarkdown.rstudio.com/) extends markdown with 
sections containing R code, and allows for notebook functionality:

* running the code in the code sections, inserting output (text, figures) in the rendered document (html, pdf)
* interacting with the document in a notebook-style (run individual code sections, scrutinize output)

These slides are written in R-markdown, and can be found as files with extension `.Rmd` on this repository: https://github.com/edzer/winterschool ; the rendered html output files are linked in the README file in the repository.

## Why is R used?

* R is a language dedicated to data analysis, unlike general programming languages like C++ or Python. In R, you don't have to load a package or module first before you can plot data, as in
 
```{r}
hist(rnorm(500))
```

* Many people use R because many other people do:
    * to solve data analysis problems
    * to share data analysis problems (share data, share R script)
    * to wrap a new data analysis method in an R package, and distribute that, e.g. through [CRAN](https://cran.r-project.org/) and/or [github](https://github.com/r-spatial/stars)


## What is an R package?

An R package wraps

* R code
* possibly further code written in C, C++, Fortran, java, etc.
* documentation
* data sets
* examples, tutorials

in a structured way, so that others can easily install it, e.g. from CRAN by

```{r eval=FALSE}
install.packages("stars")
```

or directly from github by

```{r eval=FALSE}
library(remotes) # install.packages("remotes") if you don't have this
install_github("r-spatial/stars")
```

After that, a package can be loaded by

```{r}
library(stars)
```

after which its full functionality becomes available to the R user.

**R packages extend R with new, often dedicated or domain specific functionality.**


## R-Spatial

For a historical perspective, see the [youtube video](https://www.youtube.com/watch?v=mEAyQ8bv1zU) by Roger Bivand (Why R? 2020 Keynote: "Applied Spatial Data Analysis with R: retrospect and prospect").

For a current perspective, see my [youtube video](https://www.youtube.com/watch?v=cK08bxUJn5A) (UseR! 2021 keynote, "R-Spatial").

For a current book on Spatial Data Science, see [here](https://r-spatial.org/book). We will use this book in the current summer school. Please open chapter 1.


