---
title: "Static spatial data (raster, vector); geostatistics, Machine learning for spatial prediction"
author: "Edzer Pebesma"
date: "2/16/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Geostatistics

We will work with the soil data provided by the course managers:

```{r}
library(sf)
(s = st_read("re_ca_topsoil.geojson"))
plot(s)
```

We can look at one of the variables, and try to interpolate this on a regular
grid. For this, we first need to estimate a _variogram model_ from a _sample variogram_, 
then we need a regular grid, then we need to interpolate at the grid points (kriging).

### sample variogram


Using the data as provided, we see

```{r}
library(gstat)
(v = variogram(bulk_den~1, s)) # sample variogram
plot(v)
v = variogram(bulk_den~1, s, cutoff = 40) # try with smaller cutoff
plot(v, plot.numbers = TRUE)
```


### variogram model

We choose an exponential model, and pick some initial
values for

* nugget (.1)
* range (5; this is km)
* sill (1)

that work as initial values to a non-linear fit that uses a Newton-type optimization:

```{r}
v.fit = fit.variogram(v, vgm(1, "Exp", 5, .1)) # fitted model
plot(v, v.fit) # is this O.K.?
```

### regular grid

We have target locations on a regular grid, obtained by

```{r}
library(stars)
r = read_sf("cosmoleps_ca.geojson") # grid points
plot(r, reset = FALSE)
roi = read_sf("roi_ca_geom.geojson")
plot(st_geometry(roi), add = TRUE, col = NA, border = 'orange')
```

which we can rasterize by
```{r}
ra = st_rasterize(r)
ra
plot(ra, reset = FALSE)
plot(st_geometry(roi), add = TRUE, col = NA, border = 'orange')
```

we see that cells with no points on them are assigned 0, so we need to `NA` those:
```{r}
ra[ra == 0] = NA
plot(ra, reset = FALSE)
plot(st_geometry(roi), add = TRUE, col = NA, border = 'orange')
```

### kriging

We can try to carry out kriging interpolation on this dataset, but will see that it generates warnings, mentioning singular covariance matrices, and it returns only `NA` values. A common cause of this is that we have observation locations with duplicate observations.

```{r error=TRUE, eval=FALSE}
# st_crs(ra) = st_crs(s) # -> needed for outdated R and/or packages
kr = krige(bulk_den~1, s, ra, v.fit)
```

How can we find duplicate observation locations?

```{r}
e = st_equals(s)
e
lengths(e)
which(lengths(e) > 1)
```
Removing _all_ points with an duplicate observations like we could do with


```{eval=FALSE}
s = s[lengths(e) == 1,]
```

is a bit rigorous; we'd like to _only_ remove the duplicate, not the _first_ points (or, possibly better, we'd like to compute averages for points with more than one observations). Let us remove the _duplicate_ points

```{r}
e[[11]]
e[[29]] # the second is always higher;
# take the second of each pair:
dupl = e[lengths(e) > 1]
w = sapply(dupl, tail, 1) # select the last from the duplicates
s_nd = s[-w,] # deselect those second
```

Let us try the kriging again:
```{r}
kr = krige(bulk_den~1, s_nd, ra, v.fit)
plot(kr[1], reset = FALSE, main = "bulkdensity kriging prediction")
plot(st_geometry(roi), add = TRUE, col = NA, border = 'orange')
kr$se = sqrt(kr$var1.var)
plot(kr["se"], reset = FALSE, breaks = "equal", main = "bulkdensity kriging standard error")
plot(st_geometry(roi), add = TRUE, col = NA, border = 'orange')
plot(st_geometry(s_nd), add = TRUE, col = 'orange')
```

### Doing this on a finer grid

We can create a finer (long/lat) grid, e.g. by creating a 0.01 x 0.01 **degree** raster:

```{r}
ra2 = st_as_stars(st_geometry(roi), dx = 0.01) |>
  st_crop(roi)
kr = krige(bulk_den~1, s_nd, ra2, v.fit)
plot(kr[1], reset = FALSE, main = "bulkdensity kriging prediction")
plot(st_geometry(roi), add = TRUE, col = NA, border = 'orange')
```

### Doing it properly, on a projected raster:

For instance, we can take UTM zone 33 N:

```{r}
utm33n = st_crs("EPSG:32633") # or: "+proj=utm +zone=33 +north"
roi_utm = st_transform(roi, utm33n)
s_utm = st_transform(s, utm33n)
v_utm = variogram(bulk_den~1, s_utm, cutoff = 40000) # cutoff now in m!
v_utm.fit = fit.variogram(v_utm, vgm(.01, "Exp", 15000, .02)) # fitted model
plot(v_utm, v_utm.fit) # is this O.K.?
```

We recreate a regular gird, now with grid cell size of 1 km:
```{r}
ra_utm = st_as_stars(st_geometry(roi_utm), dx = 1000) |>
  st_crop(roi_utm)
s_nd_utm = st_transform(s_nd, utm33n)
kr_utm = krige(bulk_den~1, s_nd_utm, ra_utm, v_utm.fit)
plot(kr_utm[1], reset = FALSE, main = "bulkdensity kriging prediction", axes = TRUE)
plot(st_geometry(roi_utm), add = TRUE, col = NA, border = 'orange')
```

## Machine learning

see the stars vignette: https://r-spatial.github.io/stars/articles/stars7.html

## Exercises: geostatistics

1. Load the .Rmd and carry out (reproduce) the analysis above in your R session (Rstudio)
2. Repeat the analysis for another variable from the same soil dataset
3. Try fitting a spherical variogram model, and compare this to original results
4. Look up the meaning of the arguments `cutoff` and `width` of function `variogram`, vary their values, and try to understand what they mean

Try OM (organic matter):
```{r}
v = variogram(OM~1, s_nd, cutoff = 40, width = 2) # try with smaller cutoff 
#v = variogram(OM~1, s, cutoff = 40) # try with smaller cutoff 
plot(v, plot.numbers = TRUE)
v.fit = fit.variogram(v, vgm(5, "Exp", 5, 6))
v.fit = fit.variogram(v, vgm(5, "Sph", 5, 6))
#v.fit = fit.variogram(v, vgm(5, "Exp", 5))
#v.fit = fit.variogram(v, vgm(5, "Exp", 5, 2), fit.sills = c(FALSE, TRUE))
plot(v, v.fit)
v.fit
```
