---
title: "data cube analysis of ECZ variables"
author: "Edzer Pebesma"
date: "2/17/2022"
output: html_document
---

## surface temperature

```{r}
library(stars)

load("EU_COSMOLEPS_airTemperature_mean24h.RData")
DF = DF[,-2] # remove second column, which duplicates the first
x = st_as_stars(list(DF = t(as.matrix(DF)))) |> 
	st_set_dimensions(1, st_geometry(Psf_in), name = "geometry") |>
	st_set_dimensions(2, D, name = "time")
x
```

We see that this is a 2-D data cube, with dimensions

* time
* geometry

```{r}
plot(x[,,1:9], breaks = "equal")
```

we can convert this to a multiple time series object by
```{r}
library(xts)
x.xts = as.xts(x)
plot(x.xts)
```

We can convert this to a multiple station object by
```{r}
x.sf = st_as_sf(x)
dim(x.sf)
x.sf[,1:3] # print only the first three 
plot(x.sf) # plots by default first few attributes
```

Since the points of this data cube are on a regular raster, let us
try to make a raster cube of it.

```{r}
r = st_rasterize(Psf_in)
plot(r)
```

This raster maps the points one-to-one to a raster, with zero
values where there are no points. This can be
seen from
```{r}
as.vector(r[[1]]) |> # all values in a vector
  unique() |> # all unique values
  sort() # sorted
as.vector(r[[1]]) |> # all values in a vector
  unique() |> # all unique values
  length() # includes 0 values
dim(x) # 371-1 geometries
```

Now we need to match the 370 geometries to the raster values 1-370. First we
set the zero values to `NA`:
```{r}
r[r == 0] = NA
r
```

```{r}
DF.r = t(as.matrix(DF))[as.vector(r[[1]]),]
dim(DF.r)
dim(DF.r) = c(34, 24, 255) # make it a 3D array
DF.st = st_as_stars(list(temp = DF.r))
```

We need to create a dimensions object for this matrix:
```{r}
rd = st_dimensions(r) # x&y
rd[[3]] = st_dimensions(x)[[2]] # time
names(rd)[3] = "time"
rd
st_dimensions(DF.st) = rd
plot(DF.st, breaks = "equal", col = sf.colors())
```

We can also plot fewer, e.g. every 10-th day, and add an outline of the region
```{r}
sel = seq(1, 255, by = 10)
sel
outline = st_read("roi_ca_geom.geojson")
f = function() plot(st_geometry(outline), add=TRUE, border = "orange")
plot(DF.st[,,,sel], hook = f)
```

We can do something similar with `ggplot2`:
```{r}
library(ggplot2)
ggplot() + 
		geom_stars(data = DF.st[,,,sel]) +
        facet_wrap(~time) + coord_equal() +
        theme_void() +
        scale_x_discrete(expand = c(0,0)) +
        scale_y_discrete(expand = c(0,0)) +
        scale_fill_viridis_c()
```

## Analysis

We can plot a single time series, e.g. by picking the row and column
```{r}
ts = DF.st[,20,20] |> st_as_sf(long = TRUE)
xts(ts$temp, ts$time) |> plot()
```

We can also query a (spatial) point using `st_extract()`:
```{r}
pt = st_as_sfc("POINT (14.97303 40.26846)", crs = st_crs(DF.st))
st_extract(DF.st, pt)
st_extract(DF.st, pt) |> st_as_sf(long = TRUE)
```

We can make a time slice, e.g. for the 100-th time index by
```{r}
DF.st[,,,100]
DF.st[,,,100] |> adrop()
```

We could do various aggregations, e.g. compute the mean temparture
```{r}
DF.m = st_apply(DF.st, c("x", "y"), mean)
plot(DF.m, reset = FALSE)
st_geometry(outline) |> plot(border = "orange", add = TRUE)
```

or find the day with the highest daily mean temperature
```{r}
library(dplyr) # select
mx = function(x) ifelse(all(is.na(x)), NA_real_, which.max(x))
DF.m = st_apply(DF.st, c("x", "y"), mx)
plot(DF.m, reset = FALSE)
st_geometry(outline) |> plot(border = "orange", add = TRUE)
DF.m %>%
  mutate(when = st_get_dimension_values(DF.st, "time")[.$mx]) %>%
  select(when) %>%
  plot(key.pos = 1, main = "time of maximum daily mean temp.", reset = FALSE)
st_geometry(outline) |> plot(border = "orange", add = TRUE)
```

There are only two values in the map:
```{r}
(u = unique(as.vector(DF.m[[1]])))
st_get_dimension_values(DF.st, "time")[u]
```
